### 2021-06-15 submission/manuscript.tex
Address edits from Pat

Antibiotics are a major risk factor for \emph{Clostridioides difficile}
infections (CDIs) because of their impact on the microbiota. However,
non-antibiotic medications such as the ubiquitous osmotic laxative
polyethylene glycol (PEG) 3350, also alter the microbiota, but whether
PEG impacts CDI susceptibility and clearance is unclear. To examine how
PEG impacts susceptibility, we treated C57Bl/6 mice with 5-day and 1-day
doses of 15\% PEG in the drinking water and then challenged the mice
with \emph{C. difficile} 630. We used clindamycin-treated mice as a
control because they consistently clear \emph{C. difficile} within 10
days post-challenge (dpc). PEG treatment alone was sufficient to render
mice susceptible and 5-day PEG-treated mice remain colonized for up to
30 dpc. In contrast, 1-day PEG treated mice were transiently colonized,
clearing \emph{C. difficile} within 7 dpc. To examine how PEG treatment
impacts clearance, we administered a 1-day PEG treatment to
clindamycin-treated, \emph{C. difficile}-challenged mice. Administering
PEG to mice after \emph{C. difficile} challenge prolonged colonization
up to 30 dpc. When we trained a random forest model with community data
from 5 dpc, we were able to predict which mice would exhibit prolonged
colonization (AUROC = 0.90). Examining the dynamics of these bacterial
during the post-challenge period revealed patterns in the relative
abundances of \emph{Bacteroides}, \emph{Enterobacteriaceae},
\emph{Porphyromonadaceae}, \emph{Lachnospiraceae}, and
\emph{Akkermansia} that were associated with prolonged \emph{C.
difficile} colonization in PEG-treated mice.

@@

Diarrheal samples induced by medications such as laxatives are typically
rejected for \emph{Clostridiodes difficile} testing. However, there are
some microbiota similarities between diarrheal and \emph{C. difficile}
infection (CDI) communities such as lower diversity compared to samples
from healthy patients, which lead us to hypothesize that diarrhea may be
an indicator of \emph{C. difficile} risk. We explored how osmotic
laxatives disrupt the microbiota's colonization resistance to \emph{C.
difficile} by administering a laxative to mice either before or after
\emph{C. difficile} challenge. Our findings suggest the osmotic laxative
disrupts colonization resistance to \emph{C. difficile}, as well as
clearance in mice already colonized with \emph{C. difficile}.
Considering that most hospitals recommend not performing \emph{C.
difficile} testing on patients taking laxatives and laxatives are used
when administering fecal microbiota transplants via colonoscopy to
patients with recurrent CDIs, further studies are needed to evaluate if
laxatives impact microbiota colonization resistance in humans.

@@

Many of the non-antibiotic medications associated with CDIs are known to
modulate gastrointestinal motility leading to either increased or
decreased colonic transit time, which in turn also strongly impacts
microbiota composition and function (10, 11). Stool consistency often
serves as an approximation of intestinal motility (10). Our group has
shown that when \emph{C. difficile} negative samples from patients are
separated into two groups based on stool consistency, there are shared
microbiota features such as lower alpha diversity in samples from CDI
patients and \emph{C. difficile} negative patients with diarrhea
compared to samples that were \emph{C. difficile} negative with
non-diarrheal stool consistency (12). These results led to the
hypothesis that bacterial communities from patients experiencing
diarrhea are susceptible to developing CDIs, regardless of how they
developed diarrhea.

Osmotic laxatives can lead to diarrhea depending on the dose
administered and temporarily disrupt the human intestinal microbiota
(13). The ubiquitous osmotic laxative, polyethylene glycol (PEG) 3350 is
found in Miralax, Nulytely, and Golytely and is also commonly used as
bowel preparation for colonoscopies. Interestingly, previous studies
have shown that treating mice with PEG alone altered microbiota
composition, reduced acetate and butyrate production, altered the mucus
barrier, and rendered the mice susceptible to \emph{C. difficile}
colonization (14--17). The mucus barrier is thought to mediate
protection from CDIs by protecting intestinal epithelial cells from the
toxins produced by \emph{C. difficile} (18, 19). Whether laxative
administration results in more severe CDIs in mice and how long mice
remain colonized with \emph{C. difficile} after challenge is unclear.

Beyond susceptibility, PEG is also relevant in the context of treating
recurrent CDIs via fecal microbiota transplant (FMT) where a healthy
microbiota is administered to the patient to restore colonization
resistance. For FMTs that are delivered via colonoscopy, patients
typically undergo bowel preparation by taking an osmotic laxative prior
to the procedure. Many of the FMT studies to date rationalize the use of
laxatives (20--22) based on a 1996 case study with 2 pediatric patients
where the authors suggested in the discussion that the laxative may help
flush \emph{C. difficile} spores and toxins from the intestine (23).

Our group has used C57BL/6 mice to characterize how antibiotics disrupt
the microbiota and influence \emph{C. difficile} susceptibility and
clearance (24--26). Although two groups have now shown PEG treatment
alone renders mice susceptible to \emph{C. difficile} (15, 17), these
studies have raised additional questions regarding the dynamics and
severity of infection as well as the role of laxative treatment in
\emph{C. difficile} clearance. Here, we characterized how long
PEG-treated mice remain susceptible, whether PEG treatment results in
more sustained \emph{C. difficile} colonization and severe CDI than mice
treated with clindamycin, and if PEG treatment after challenge can
promote \emph{C. difficile} clearance. Addressing these questions will
better inform how we think about laxatives and diarrhea in the context
of CDIs.

@@

\textbf{5-day laxative treatment led to prolonged \emph{C. difficile}
colonization in mice.} Building off of previous work that showed
treating mice with the osmotic laxative, PEG 3350, rendered mice
susceptible to \emph{C. difficile} colonization (15, 17), we decided to
test how long \emph{C. difficile} colonization is sustained and how long
PEG-treated mice remain susceptible to \emph{C. difficile}. We compared
three groups of mice treated with PEG 3350 to one group of mice treated
with our standard 10 mg/kg clindamycin treatment, which temporarily
renders the mice susceptible to \emph{C. difficile} colonization, with
mice typically clearing \emph{C. difficile} within 10 days
post-challenge (9, 26). All three groups of PEG-treated mice were
administered a 15\% PEG solution in the drinking water for 5-days. The
first group received no additional treatment. The second group was also
treated with clindamycin. A third group was allowed to recover for 10
days prior to challenge (Fig. 1A). The PEG treatment resulted in weight
loss for the 3 groups of mice, with the greatest change in weight
observed on the fifth day of the PEG treatment. The mice recovered most
of the lost weight by five days after treatment (Fig. 1B). After either
the PEG, clindamycin, or PEG and clindamycin treatment all mice were
challenged with 10\textsuperscript{5} \emph{C. difficile} 630 spores
(Fig. 1A). All treatments rendered mice susceptible to \emph{C.
difficile} colonization. In contrast to the mice that only received
clindamycin, PEG-treated mice remained colonized with \emph{C.
difficile} at a high level through thirty days post-challenge (Fig. 1C).
The clindamycin-treated mice cleared \emph{C. difficile} within ten days
post-challenge (Fig. 1C). It was noteworthy that PEG-treated mice were
still susceptible to \emph{C. difficile} colonization after a 10-day
recovery period, although \emph{C. difficile} was not detectable in most
of the group in the initial five days post-challenge (Fig. 1C, S1A). One
mouse was found dead on the 6th day post-challenge, presumably due to
\emph{C. difficile} as the bacterium became detectable in stool samples
from that mouse on the 4th day post-challenge (Fig. S1A, mouse 10). From
8 days post-challenge onward, the density of \emph{C. difficile}
stabilized in the 10-day recovery group and remained high through 20-30
days post-challenge (Fig. 1C). Thus, osmotic laxative treatment alone
was sufficient to render mice susceptible to prolonged \emph{C.
difficile} colonization and PEG-treated mice remained susceptible
through ten days post-treatment.

@@

Since none of the communities completely recovered in the follow-up
period after treatments, we next profiled community diversity and
composition. We examined the alpha diversity dynamics by calculating the
communities' Shannon diversity. Although both clindamycin and PEG
treatments decreased diversity, Shannon diversity was lower in the
groups of mice that received PEG treatment compared to those that
received clindamycin through thirty days post-challenge (Fig. 2C; Data
Set S1, sheet 2). We next examined the bacterial genera that shifted
after PEG treatment by comparing the baseline samples of mice treated
with only PEG to samples from the same mice one day post-PEG-treatment.
We found 18 genera that were altered by PEG treatment (Data Set S1,
sheet 3). The majority of the bacterial relative abundances decreased
after the PEG treatment, but the relative abundance among members of the
\emph{Enterobacteriaceae} and \emph{Bacteroides} increased. The increase
in \emph{Bacteroides} relative abundance was unique to PEG treated mice,
as the \emph{Bacteroides} relative abundance actually decreased in
clindamycin treated mice (Fig. 2D). Finally, we identified the genera
whose relative abundance differed across treatment groups over multiple
time points. Of the 33 genera that were different between treatment
groups, 24 genera were different over multiple time points (Fig. 2E,
Data Set S1, sheet 4). Thus, PEG had a significant impact on the fecal
microbiota that was maintained over time and was distinct from
clindamycin treatment.

@@

Because there were differences in the mucosal microbiota including
detectable \emph{C. difficile} sequences in tissues from PEG-treated
mice relative to mice treated with clindamycin, we next examined the
severity of \emph{C. difficile} challenge by evaluating cecum and colon
histopathology (27). However, we found there was no difference in cecum
and colon scores between clindamycin and PEG-treated mice that were
challenged with \emph{C. difficile} at 4 days post-challenge (Fig. 3D),
the time point typically examined in \emph{C. difficile} 630 challenged
mice (28). We also looked at 6 days post-challenge because that was when
there was a large difference in \emph{C. difficile} density between PEG-
and clindamycin-treated mice (Fig. 1C). Although there was a slight
difference in the histopathology score of the colon between PEG and
clindamycin-treated mice, there was not a signifant difference in the
cecum and the overall score was relatively low (1.5 to 2.5 out of 12,
Fig. 3E). Therefore, although PEG treatment had a disruptive effect on
the mucosal microbiota, the impact of \emph{C. difficile} 630 challenge
on the cecum and colon was similar between PEG and clindamycin treated
mice.

@@

\textbf{Five-day post-challenge community data can predict which mice
will have prolonged \emph{C. difficile} colonization.} After identifying
bacteria associated with the 5-day, 1-day and post-challenge 1-day PEG
treatments, we examined the bacteria that influenced prolonged \emph{C.
difficile} colonization. We trained 3 machine learning models (random
forest, logistic regression, and support vector machine) with bacterial
community data from 5 days post-challenge to predict whether the mice
were still colonized with \emph{C. difficile} 10 days post-challenge. We
chose 5 days post-challenge because that was the earliest time point
where we saw a treatment effect in the post-challenge 1-day PEG
experiments. The random forest model had the highest performance (median
AUROC = 0.90, Data Set S1, sheet 19), suggesting the 5-day post
challenge microbiota is an excellent predictor of prolonged \emph{C.
difficile} colonization. Next, we performed permutation importance to
examine the bacteria that were the top contributors to the random forest
model predicting prolonged \emph{C. difficile }colonization. We selected
10 genera that contributed the most to our model's performance (Fig. 7A)
and examined their relative abundance at 5 days post-challenge, the time
point used to predict \emph{C. difficile} colonization status on day 10
(Fig. 7B). Next, we focused on the 5 genera that had a greater than 1\%
relative abundance in either the cleared or colonized mice and examined
how the bacteria changed over time. We found \emph{Enterobacteriaceae}
and \emph{Bacteroides} tended to have a higher relative abundance, the
relative abundance of \emph{Akkermansia} was initially decreased and
then increased, and \emph{Porphyromonadaceae} and \emph{Lachnospiraceae}
had a lower relative abundance in the mice with prolonged colonization
compared to the mice that cleared \emph{C. difficile} (Fig. 7C).
Together these results suggest a combination of low and high abundance
bacterial genera influence the prolonged colonization observed in 5-day
PEG and post-challenge 1-day PEG treated mice.

@@

While the disruptive effect of antibiotics on \emph{C. difficile}
colonization resistance is well established, the extent to which other
drugs such as laxatives disrupt colonization resistance was unclear. By
following osmotic laxative treated mice over time, we found that a 5-day
PEG treatment before challenge resulted in prolonged \emph{C. difficile}
colonization, while a 1-day PEG treatment resulted in transient
colonization. The differences in \emph{C. difficile} colonization
dynamics between the 5- and 1-day PEG treated mice were associated with
differences in how much the treatments disrupted the microbiota.
Additionally, the intestinal communities of 5-day PEG treated mice did
not regain colonization resistance after a 10-day recovery period. In
contrast to the other 5-day PEG treatment groups, \emph{C. difficile}
was not immediately detectable in the stools of most of the mice in the
10-day recovery group. We also examined the impact of PEG treatment
after \emph{C. difficile} challenge. In opposition to the hypothesis
suggested by the literature, we found that PEG treatment prolonged
colonization relative to mice that only recieved clindamycin treatment.
We identified patterns in the relative abundances of \emph{Bacteroides},
\emph{Enterobacteriaceae}, \emph{Akkermansia},
\emph{Porphyromonadaceae}, and \emph{Lachnospiraceae} that were
associated with prolonged \emph{C. difficile} colonization (Fig. 8).
Overall, our results demonstrated that osmotic laxative treatment alone
rendered mice susceptible to \emph{C. difficile} colonization and the
duration of colonization depended on the length of PEG treatment and
whether treatment was administered before or after challenge.

@@

Considering laxatives are also used to prepare patients when
administering fecal microbiota transplants via colonoscopy to treat
recurrent CDIs, it will be important to determine if osmotic laxatives
impact \emph{C. difficile} clearance in the human intestinal tract. It
is still unclear what the best administration route is because there
have been no studies designed to evaluate the best administration route
for FMTs (68). Nevertheless, results from the FMT National Registry
where 85\% of FMTs were delivered by colonoscopy demonstrate FMTs are
highly effective treatment for recurrent CDIs with 90\% achieving
resolution in the 1 month follow-up window (69). A surprising number of
studies continue to hypothesize that PEG or bowel preparation can clear
\emph{C. difficile} spores and toxins despite the paucity of supporting
evidence (20--23). There was even a clinical trial (NCT01630096)
designed to examine whether administering PEG 3350 (NuLYTELY) prior to
antibiotic treatment reduced disease severity that started recruitment
in 2012 (70), but no results have been posted to date. Here we sought to
evaluate the impact of treating \emph{C. difficile} colonized mice with
PEG (with or without FMT) and found clearance was delayed. Further
studies are needed to understand the impact of osmotic laxatives on
\emph{C. difficile} colonization resistance and clearance in human
patients receiving FMTs.

@@

\textbf{\emph{C. difficile} challenge model.} Mice were challenged with
25 microliters off \emph{C. difficile} 630 spores at
10\textsuperscript{5} concentration, except for 1 experiment where the
concentration was 10\textsuperscript{3} (Fig. 5A). All mock challenged
mice received 25 ul vehicle solution (Ultrapure water). A Dymax stepper
pipette was used to administer the same challenge dose to mice via oral
gavage. Mice were weighed daily throughout the experiment and stool was
collected for quantifying \emph{C. difficile} CFU and 16S rRNA gene
sequencing. There were two groups of mice that received either a PBS or
fecal microbiota transplant (FMT) gavage post-PEG treatment. The fecal
microbiota transplant was prepared with stool samples collected from the
mice in the experiment prior to the start of any treatments. The stool
samples were transferred to an anaerobic chamber and diluted 1:10 in
reduced PBS and glycerol was added to make a 15\% glycerol solution. The
solution was then aliquoted into tubes and stored at -80°C until the day
of the gavage. An aliquot of both the FMT and PBS solutions were also
set aside in the -80°C for 16S rRNA gene sequencing. The day of the
gavage, aliquots were thawed and centrifuged at 7500 RPM for 1 minute.
The supernatant was then transferred to a separate tube to prevent the
gavage needle from clogging with debris during gavage. The PBS solution
that was administered to the other group was also 15\% glycerol. Each
mouse was administered 100 microliters of either the FMT or PBS solution
via gavage. When we refer to mice that cleared \emph{C. difficile}, we
mean that no \emph{C. difficile} was detected in the first serial
dilution (limit of detection: 100 CFU). In some experiments, we
collected tissues for 16SrRNA gene sequencing, histopathology, or both.
For 16S rRNA gene sequencing, we collected small snips of cecum,
proximal colon, and distal colon tissues in microcentrifuge tubes, snap
froze in liquid nitrogen, and stored at -80°C. For histopathology, cecum
and colon tissues were placed into separate casettes, fixed, and then
submitted to McClinchey Histology Labs (Stockbridge, MI) for processing,
embedding, and hematoxylin and eosin (H\&E) staining.

@@

\textbf{16S rRNA gene sequencing.} Stool samples that were stored in the
-80°C were placed into 96-well plates for DNA extractions and library
preparation. DNA was extracted using the DNeasy Powersoil HTP 96 kit
(Qiagen) and an EpMotion 5075 automated pipetting system (Eppendorf).
For library preparation, each plate had a mock community control
(ZymoBIOMICS microbial community DNA standards) and a negative control
(water). The V4 region of the 16S rRNA gene was amplified with the
AccuPrime Pfx DNA polymerase (Thermo Fisher Scientific) using custom
barcoded primers, as previously described (71). The PCR amplicons were
normalized (SequalPrep normalizatin plate kit from Thermo Fisher
Scientific), pooled and quantified (KAPA library quantification kit from
KAPA Biosystems), and sequenced with the MiSeq system (Illumina).

>>>>

Antibiotics are a major risk factor for \emph{Clostridioides difficile}
infections (CDIs) because of their impact on the microbiota. However,
non-antibiotic medications such as the ubiquitous osmotic laxative
polyethylene glycol (PEG) 3350 also alter the microbiota. But whether
PEG impacts CDI susceptibility and clearance is unclear. To examine how
PEG impacts susceptibility, we treated C57Bl/6 mice with 5-day and 1-day
doses of 15\% PEG in the drinking water and then challenged the mice
with \emph{C. difficile} 630. We used clindamycin-treated mice as a
control because they consistently clear \emph{C. difficile} within 10
days post-challenge. PEG treatment alone was sufficient to render mice
susceptible and 5-day PEG-treated mice remained colonized for up to 30
days post-challenge. In contrast, 1-day PEG treated mice were
transiently colonized, clearing \emph{C. difficile} within 7 days
post-challenge. To examine how PEG treatment impacts clearance, we
administered a 1-day PEG treatment to clindamycin-treated, \emph{C.
difficile}-challenged mice. Administering PEG to mice after \emph{C.
difficile} challenge prolonged colonization up to 30 days
post-challenge. When we trained a random forest model with community
data from 5 days post-challenge, we were able to predict which mice
would exhibit prolonged colonization (AUROC = 0.90). Examining the
dynamics of these bacterial populations during the post-challenge period
revealed patterns in the relative abundances of \emph{Bacteroides},
\emph{Enterobacteriaceae}, \emph{Porphyromonadaceae},
\emph{Lachnospiraceae}, and \emph{Akkermansia} that were associated with
prolonged \emph{C. difficile} colonization in PEG-treated mice. Thus,
the osmotic laxative, PEG, rendered mice susceptible to \emph{C.
difficile} colonization and hindered clearance.

@@

Diarrheal samples from patients taking laxatives are typically rejected
for \emph{Clostridiodes difficile} testing. However, there are
similarities between the bacterial communities from people with diarrhea
or \emph{C. difficile} infections (CDI) including lower diversity
compared to communities from healthy patients, which led us to
hypothesize that diarrhea may be an indicator of \emph{C. difficile}
risk. We explored how osmotic laxatives disrupt the microbiota's
colonization resistance to \emph{C. difficile} by administering a
laxative to mice either before or after \emph{C. difficile} challenge.
Our findings suggest that osmotic laxatives disrupt colonization
resistance to \emph{C. difficile}, and prevent clearance among mice
already colonized with \emph{C. difficile}. Considering that most
hospitals recommend not performing \emph{C. difficile} testing on
patients taking laxatives and laxatives are used when administering
fecal microbiota transplants via colonoscopy to patients with recurrent
CDIs, further studies are needed to evaluate if laxatives impact
microbiota colonization resistance in humans.

@@

Many of the non-antibiotic medications associated with CDIs are known to
modulate gastrointestinal motility leading to either increased or
decreased colonic transit time, which in turn also strongly impacts
microbiota composition and function (10, 11). Stool consistency often
serves as an approximation of intestinal motility (10). Our group has
shown that when \emph{C. difficile} negative samples from patients are
separated into two groups based on stool consistency, there are similar
microbiota features between samples from CDI patients and \emph{C.
difficile} negative patients with diarrhea compared to samples that were
\emph{C. difficile} negative with non-diarrheal stool consistency (12).
The similar community features between CDI patients and patients with
diarrhea included low alpha diversity and only 6 bacterial taxa had
higher relative abundances in communities from CDI patients. These
results led to the hypothesis that bacterial communities from patients
experiencing diarrhea are susceptible to developing CDIs, regardless of
how they developed diarrhea.

depending on the dose administered, osmotic laxatives can lead to
diarrhea and temporarily disrupt the human intestinal microbiota (13).
The ubiquitous osmotic laxative, polyethylene glycol (PEG) 3350 is found
in Miralax, Nulytely, and Golytely and is also commonly used as bowel
preparation for colonoscopies. Interestingly, previous studies have
shown that treating mice with PEG alone altered microbiota composition,
reduced acetate and butyrate production, altered the mucus barrier, and
rendered the mice susceptible to \emph{C. difficile} colonization
(14--17). The mucus barrier is thought to mediate protection from CDIs
by protecting intestinal epithelial cells from the toxins produced by
\emph{C. difficile} (18, 19). Whether laxative administration results in
more severe CDIs in mice and how long mice remain colonized with
\emph{C. difficile} after challenge is unclear.

Beyond susceptibility, PEG is also relevant in the context of treating
recurrent CDIs via fecal microbiota transplant (FMT) where a healthy
microbiota is administered to the patient to restore colonization
resistance. For FMTs that are delivered via colonoscopy, patients
typically undergo bowel preparation by taking an osmotic laxative prior
to the procedure. Many of the FMT studies to date rationalize the use of
laxatives prior to the FMT (20--22) based on a 1996 case study with 2
pediatric patients where the authors suggested in the discussion that
the laxative may help flush \emph{C. difficile} spores and toxins from
the intestine (23).

Our group has used C57BL/6 mice to characterize how antibiotics disrupt
the microbiota and influence \emph{C. difficile} susceptibility and
clearance (24--26). Although two groups have now shown that PEG
treatment alone renders mice susceptible to \emph{C. difficile} (15,
17), these studies have raised additional questions regarding the
dynamics and severity of infection as well as the role of laxative
treatment in \emph{C. difficile} clearance. Here, we characterized how
long PEG-treated mice remain susceptible, whether PEG treatment results
in more sustained \emph{C. difficile} colonization and severe CDI than
mice treated with clindamycin, and whether PEG treatment after challenge
can promote \emph{C. difficile} clearance. Addressing these questions
will better inform how we think about laxatives and diarrhea in the
context of CDIs.

@@

\textbf{5-day laxative treatment led to prolonged \emph{C. difficile}
colonization in mice.} Building off of previous work that showed
treating mice with the osmotic laxative, PEG 3350, rendered mice
susceptible to \emph{C. difficile} colonization (15, 17), we decided to
test how long \emph{C. difficile} colonization is sustained and how long
PEG-treated mice remain susceptible to \emph{C. difficile}. We compared
three groups of mice treated with PEG 3350 to one group of mice treated
with our standard 10 mg/kg clindamycin treatment, which temporarily
renders mice susceptible to \emph{C. difficile} colonization, with mice
typically clearing \emph{C. difficile} within 10 days post-challenge (9,
26). All three groups of PEG-treated mice were administered a 15\% PEG
solution in the drinking water for 5-days. The first group received no
additional treatment. The second group was also treated with
clindamycin. A third group was allowed to recover for 10 days prior to
challenge (Fig. 1A). The PEG treatment resulted in weight loss for the 3
groups of mice, with the greatest change in weight observed on the fifth
day of the PEG treatment. The mice recovered most of the lost weight by
five days after treatment (Fig. 1B). After either the PEG, clindamycin,
or PEG and clindamycin treatment all mice were challenged with
10\textsuperscript{5} \emph{C. difficile} 630 spores (Fig. 1A). All
treatments rendered mice susceptible to \emph{C. difficile}
colonization. In contrast to the mice that only received clindamycin,
PEG-treated mice remained colonized with \emph{C. difficile} at a high
level through thirty days post-challenge (Fig. 1C). The
clindamycin-treated mice cleared \emph{C. difficile} within ten days
post-challenge (Fig. 1C). It was noteworthy that PEG-treated mice were
still susceptible to \emph{C. difficile} colonization after a 10-day
recovery period, although \emph{C. difficile} was not detectable in most
of the group in the initial five days post-challenge (Fig. 1C, S1A). One
mouse was found dead on the 6th day post-challenge, presumably due to
\emph{C. difficile} as the bacterium became detectable in stool samples
from that mouse on the 4th day post-challenge (Fig. S1A, mouse 10). From
8 days post-challenge onward, the density of \emph{C. difficile}
stabilized in the 10-day recovery group and remained high through 20-30
days post-challenge (Fig. 1C). Thus, osmotic laxative treatment alone
was sufficient to render mice susceptible to prolonged \emph{C.
difficile} colonization and PEG-treated mice remained susceptible
through ten days post-treatment.

@@

Because time and treatment group influenced most of the variation
between communities, we next profiled community diversity and
composition. We examined the alpha diversity dynamics by calculating the
communities' Shannon diversity. Although both clindamycin and PEG
treatments decreased diversity, the Shannon diversity was lower in the
groups of mice that received PEG treatment compared to those that
received clindamycin alone through thirty days post-challenge (Fig. 2C;
Data Set S1, sheet 2). We next identified the bacterial genera whose
relative abundances shifted after PEG treatment by comparing the
baseline samples of mice treated with only PEG to samples from the same
mice one day post-PEG-treatment. We found 18 genera whose relative
abundances were altered by PEG treatment (Data Set S1, sheet 3). The
majority of the bacterial relative abundances decreased after the PEG
treatment, but the relative abundance among members of the
\emph{Enterobacteriaceae} and \emph{Bacteroides} increased. The increase
in \emph{Bacteroides} relative abundance was unique to PEG treated mice,
as the \emph{Bacteroides} relative abundance actually decreased in
clindamycin treated mice (Fig. 2D). Finally, we identified the genera
whose relative abundance differed across treatment groups over multiple
time points. Of the 33 genera that were different between treatment
groups, 24 genera were different over multiple time points (Fig. 2E,
Data Set S1, sheet 4). Thus, PEG had a significant impact on the fecal
microbiota that was maintained over time and was distinct from
clindamycin treatment.

@@

Because there were differences in the mucosal microbiota including
detectable \emph{C. difficile} sequences in tissues from PEG-treated
mice relative to mice treated with clindamycin, we next examined the
severity of \emph{C. difficile} challenge by evaluating cecum and colon
histopathology (27). However, we found there was no difference in cecum
and colon scores between clindamycin and PEG-treated mice that were
challenged with \emph{C. difficile} at 4 days post-challenge (Fig. 3D),
the time point typically examined in \emph{C. difficile} 630 challenged
mice (28). We also looked at 6 days post-challenge because that was when
there was a large difference in \emph{C. difficile} density between PEG-
and clindamycin-treated mice (Fig. 1C). Although there was a slight
difference in the histopathology score of the colon between PEG and
clindamycin-treated mice, there was not a signifant difference in the
cecum and the overall score was relatively low (1.5 to 2.5 out of 12,
Fig. 3E). Therefore, although PEG treatment had a disruptive effect on
the mucosal microbiota, the impact of \emph{C. difficile} challenge on
the cecum and colon was similar between PEG and clindamycin treated
mice.

@@

\textbf{Five-day post-challenge community data can predict which mice
will have prolonged \emph{C. difficile} colonization.} After identifying
bacteria associated with the 5-day, 1-day and post-challenge 1-day PEG
treatments, we examined the bacteria that influenced prolonged \emph{C.
difficile} colonization. We trained 3 machine learning models (random
forest, logistic regression, and support vector machine) with bacterial
community data from 5 days post-challenge to predict whether the mice
were still colonized with \emph{C. difficile} 10 days post-challenge. We
chose to predict the status based on communities 5 days post-challenge
because that was the earliest time point where we saw a treatment effect
in the post-challenge 1-day PEG experiments. The random forest model had
the highest performance (median AUROC = 0.90, Data Set S1, sheet 19) and
indicated that the 5-day post challenge microbiota was an excellent
predictor of prolonged \emph{C. difficile} colonization. Next, we
performed a permutation importance test to identify the bacteria that
were the top contributors to the random forest model for predicting
prolonged \emph{C. difficile }colonization. We selected 10 genera that
contributed the most to our model's performance (Fig. 7A) and examined
their relative abundance at 5 days post-challenge, the time point used
to predict \emph{C. difficile} colonization status on day 10 (Fig. 7B).
Next, we focused on the 5 genera that had a greater than 1\% relative
abundance in either the cleared or colonized mice and examined how the
bacteria changed over time. We found \emph{Enterobacteriaceae} and
\emph{Bacteroides} tended to consistently have a higher relative
abundance, the relative abundance of \emph{Akkermansia} was initially
low and then increased, and \emph{Porphyromonadaceae} and
\emph{Lachnospiraceae} had a lower relative abundance in the mice with
prolonged colonization compared to the mice that cleared \emph{C.
difficile} (Fig. 7C). Together these results suggest a combination of
low and high abundance bacterial genera influence the prolonged
colonization observed in 5-day PEG and post-challenge 1-day PEG treated
mice.

@@

While the disruptive effect of antibiotics on \emph{C. difficile}
colonization resistance is well established, the extent to which other
drugs such as laxatives disrupt colonization resistance was unclear. By
following mice treated with an osmotic laxative over time, we found that
a 5-day PEG treatment before challenge resulted in prolonged \emph{C.
difficile} colonization, while a 1-day PEG treatment resulted in
transient colonization without the use of antibiotics. The differences
in \emph{C. difficile} colonization dynamics between the 5- and 1-day
PEG treated mice were associated with differences in how much the
treatments disrupted the microbiota. Additionally, the intestinal
communities of 5-day PEG treated mice did not regain colonization
resistance after a 10-day recovery period. In contrast to the other
5-day PEG treatment groups, \emph{C. difficile} was not immediately
detectable in the stools of most of the mice in the 10-day recovery
group. We also examined the impact of PEG treatment after \emph{C.
difficile} challenge. In opposition to the hypothesis suggested by the
literature, we found that PEG treatment prolonged colonization relative
to mice that only recieved clindamycin treatment. We identified patterns
in the relative abundances of \emph{Bacteroides},
\emph{Enterobacteriaceae}, \emph{Akkermansia},
\emph{Porphyromonadaceae}, and \emph{Lachnospiraceae} that were
associated with prolonged \emph{C. difficile} colonization (Fig. 8).
Overall, our results demonstrated that osmotic laxative treatment alone
rendered mice susceptible to \emph{C. difficile} colonization and the
duration of colonization depended on the length of PEG treatment and
whether treatment was administered before or after challenge.

@@

Considering laxatives are also used to prepare patients when
administering fecal microbiota transplants via colonoscopy to treat
recurrent CDIs, it will be important to determine whether osmotic
laxatives impact \emph{C. difficile} clearance in the human intestinal
tract. It is still unclear what the best administration route is because
there have been no studies designed to evaluate the best administration
route for FMTs (68). Nevertheless, results from the FMT National
Registry where 85\% of FMTs were delivered by colonoscopy demonstrate
FMTs are highly effective treatment for recurrent CDIs with 90\%
achieving resolution in the 1 month follow-up window (69). A surprising
number of studies continue to hypothesize that PEG or bowel preparation
can clear \emph{C. difficile} spores and toxins despite the paucity of
supporting evidence (20--23). There was even a clinical trial
(NCT01630096) designed to examine whether administering PEG 3350
(NuLYTELY) prior to antibiotic treatment reduced disease severity that
started recruitment in 2012 (70), but no results have been posted to
date. Here we sought to evaluate the impact of treating \emph{C.
difficile} colonized mice with PEG (with or without FMT) and found
clearance was delayed. Further studies are needed to understand the
impact of osmotic laxatives on \emph{C. difficile} colonization
resistance and clearance in human patients receiving FMTs.

@@

\textbf{\emph{C. difficile} challenge model.} Mice were challenged with
25 microliters of \emph{C. difficile} 630 spores at
10\textsuperscript{5} concentration, except for 1 experiment where the
concentration was 10\textsuperscript{3} (Fig. 5A). All mock challenged
mice received 25 ul vehicle solution (Ultrapure water). A Dymax stepper
pipette was used to administer the same challenge dose to mice via oral
gavage. Mice were weighed daily throughout the experiment and stool was
collected for quantifying \emph{C. difficile} CFU and 16S rRNA gene
sequencing. There were two groups of mice that received either a PBS or
fecal microbiota transplant (FMT) gavage post-PEG treatment. The fecal
microbiota transplant was prepared with stool samples collected from the
mice in the experiment prior to the start of any treatments. The stool
samples were transferred to an anaerobic chamber and diluted 1:10 in
reduced PBS and glycerol was added to make a 15\% glycerol solution. The
solution was then aliquoted into tubes and stored at -80°C until the day
of the gavage. An aliquot of both the FMT and PBS solutions were also
set aside in the -80°C for 16S rRNA gene sequencing. The day of the
gavage, aliquots were thawed and centrifuged at 7500 RPM for 1 minute.
The supernatant was then transferred to a separate tube to prevent the
gavage needle from clogging with debris during gavage. The PBS solution
that was administered to the other group was also 15\% glycerol. Each
mouse was administered 100 microliters of either the FMT or PBS solution
via gavage. When we refer to mice that cleared \emph{C. difficile}, we
mean that no \emph{C. difficile} was detected in the first serial
dilution (limit of detection: 100 CFU). In some experiments, we
collected tissues for 16SrRNA gene sequencing, histopathology, or both.
For 16S rRNA gene sequencing, we collected small snips of cecum,
proximal colon, and distal colon tissues in microcentrifuge tubes, snap
froze in liquid nitrogen, and stored at -80°C. For histopathology, cecum
and colon tissues were placed into separate casettes, fixed, and then
submitted to McClinchey Histology Labs (Stockbridge, MI) for processing,
embedding, and hematoxylin and eosin (H\&E) staining.

@@

\textbf{16S rRNA gene sequencing.} Stool samples were stored at -80°C
and were placed into 96-well plates for DNA extractions and library
preparation. DNA was extracted using the DNeasy Powersoil HTP 96 kit
(Qiagen) and an EpMotion 5075 automated pipetting system (Eppendorf).
For library preparation, each plate had a mock community control
(ZymoBIOMICS microbial community DNA standards) and a negative control
(water). The V4 region of the 16S rRNA gene was amplified with the
AccuPrime Pfx DNA polymerase (Thermo Fisher Scientific) using custom
barcoded primers, as previously described (71). The PCR amplicons were
normalized (SequalPrep normalizatin plate kit from Thermo Fisher
Scientific), pooled and quantified (KAPA library quantification kit from
KAPA Biosystems), and sequenced with the MiSeq system (Illumina).

\textbf{16S rRNA gene sequence analysis.} All sequences were processed
with mothur (v. 1.43) using a previously published protocol (71, 72).
Paired sequencing reads were combined and aligned with the SILVA (v.
132) reference database (73) and taxonomy was assigned with a modified
version of the Ribosomal Database Project reference sequences (v. 16)
(74). The error rate for are sequencing data was 0.0559\% based on the
17 mock communities we ran with the samples. Samples were rarefied to
1,000 sequences, 1,000 times for alpha and beta diversity analyses in
order to account for uneven sequencing across samples. All but 3 of the
17 water controls had fewer than 1000 sequences. PCoAs were generated
based on Bray-Curtis Index distance matrices. Permutational multivariate
analysis of variance (PERMANOVA) tests were performed on
mothur-generated Bray-Curtis distance matrices with the adonis function
from the vegan R package (75).
